### 2014-11-12 intro.tex
Fixed several typos.

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race conditions. Such
issues are handled with low level constructs such as locks and condition
variables, requiring a fair amount of effort to get right.  Declarative
programming has been hailed as a solution to this issue, since the problem of
implementing the details of parallelism is moved from the programmer to the
compiler and runtime. The programmer writes code without having to deal with
parallel programming constructs and the compiler automatically parallelizes the
program in order to take advantage of multiple threads of execution.  The
programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although general declarative languages have yet to be as successful, the future
looks promising for this particular approach.

The problem with declarative programming is that it leaves little to no
programmer control over how execution is scheduled or how data is laid out,
making it hard to achieve necessary efficiency. This introduces
performance issues because even if the runtime system is able to
reasonably parallelize the program using a general algorithm, there
is a lack of specific information about the program that a compiler
cannot easily deduce. Such information could make execution better in
terms of run time time, memory usage, and scalability.

In this paper, we introduce Coordinated Linear Meld (CLM), a data-centric declarative
language that extends the Linear Meld~(LM)
language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give
programmer control over scheduling and data placement. LM is a linear logic
programming language designed for programs that operate on graphs.  The use
of linear logic~\cite{girard-87} supports structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to \scare{communicate} with other nodes.  Both
computation and communication happen through the derivation of logical rules
(which make up the program).

The CLM language features coordination primitives that can be used in the same
way as any other primitive, in other words, they are specified with the same
syntax and semantics as the rest of the programming language. These coordination
primitives can be used to improve program execution based on the state of the
program and the underlying machine. The first kind of coordination facts are
called \emph{sensing facts} and are used to sense information about the system
the program is running on, e.g., scheduling and node placement on threads. The
second kind of coordination facts are \emph{action facts} that when deleted in
rules are used to apply a scheduling operation during execution. Coordination
facts allow the programmer to write logical rules that depend on the current
state of the program and then prioritize node computation or place nodes in
different threads.

To the best of our knowledge, this is the first time that a declarative language
allows control over execution while staying declarative and without resorting to
meta-language constructs. This is crucial to our goal of being able to prove
programs correct since proofs can be constructed even in the presence of
coordination facts.  After briefly discussing related work, we present an
overview of the base language, with an example. In
\sectref{sec:coordination}, we introduce the current set of coordination
facts, followed by a description of the changes required to implement the
desired coordination mechanisms. In \sectref{sec:applications} we present
several applications and show coordination can improve programs without
destroying clarity or provability.

>>>>

Parallel programming in sequential languages is hard because manipulating a
shared state using multiple threads may result in data race conditions. Such
issues are handled with low level constructs such as locks, semaphores and/or condition
variables, requiring a fair amount of effort to get right.  Declarative
programming has been hailed as a alternative solution to this issue, since the problem of
implementing the details of parallelism is moved from the programmer to the
compiler and runtime environment. The programmer writes code without having to deal with
parallel programming constructs and the compiler automatically parallelizes the
program in order to take advantage of multiple threads of execution.  This
programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although general declarative languages have yet to be as successful, the future
looks promising for this particular approach.

The problem with declarative programming is that it leaves little to no
programmer control over how execution is scheduled or how data is laid out,
making it hard to improve efficiency. This introduces
performance issues because even if the runtime system is able to
reasonably parallelize the program using a general algorithm, there
is a lack of specific information about the program that a compiler
cannot easily deduce. Such information could make execution better in
terms of run time time, memory usage, and/or scalability.

In this paper, we introduce Coordinated Linear Meld (CLM), a data-centric declarative
language that extends the Linear Meld~(LM)
language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give
programmer control over scheduling and data placement. LM is a linear logic
programming language designed for programs that operate on graphs.  The use
of linear logic~\cite{girard-87} supports structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently but is allowed to \scare{communicate} with other nodes.  Both
computation and communication happen through the derivation of logical rules
(which make up the program).

The CLM language features two kinds of coordination primitives that can be used in the same
way as any other primitive, i.e., they are specified with the same
syntax and semantics as the rest of the programming language. These coordination
primitives can be used to improve program execution based on the state of the
program and the underlying machine. The first kind of coordination facts are
called \emph{sensing facts} and are used to sense information about the system
the program is running on, e.g., scheduling and node placement on threads. The
second kind of coordination facts are \emph{action facts} that when detected in
rules, are used to apply a scheduling operation during execution. Coordination
facts allow the programmer to write logical rules that depend on the current
state of the program and then prioritize node computation or place nodes in
different threads.

To the best of our knowledge, this is the first time that a declarative language
allows control over execution while staying declarative and without resorting to
meta-language constructs. This is crucial to our goal of being able to prove
programs correct since proofs can be constructed even in the presence of
coordination facts.  After briefly discussing related work, we present an
overview of the base language, with an example. In
\sectref{sec:coordination}, we introduce the current set of coordination
facts, followed by a description of the changes required to implement the
desired coordination mechanisms. In \sectref{sec:applications} we present
several applications and show how coordination can improve programs without
destroying clarity or provability.

### 2014-11-12 related.tex
Fixed several typos.

\paragraph{Declarative Languages}
Declarative programming paradigms such as \emph{logic} and \emph{functional}
programming have been extensively exploited for their implicit parallelism.
In logic languages such as Prolog, researchers took advantage of the
non-determinism of proof-search to evaluate subgoals in parallel with models
such as \emph{or-parallelism}~\cite{ali-86} and
\emph{and-parallelism}~\cite{Shen-92}.  In functional languages, the stateless
nature of computation allows multiple expressions to evaluate safely in
parallel. This has been explored in several languages such as
NESL~\cite{Blelloch:1996:PPA:227234.227246}, Id~\cite{Nikhil93anoverview}, and
more recently Data Parallel Haskell~\cite{nepal}.
NESL often got good performance, but was limited to a particular
application space. We also limit our application space, but also give the
programmer a declarative way to coordinate computation to improve the
performance of the program.

@@

GraphLab~\cite{GraphLab2010} is a C++ framework for developing graph-based parallel machine
learning algorithms. GraphLab allows nodes to have read/write access to
different scopes through different concurrent access models in order to balance
performance and data consistency.  GraphLab also provides different schedulers
that dictate the order in which node's are computed, allowing the programmer to
optimize the program. Later in this paper, we will show how one GraphLab
scheduler can be implemented in CLM through the use of coordination facts.
Another, more restrictive, graph-based system is
Pregel~\cite{Malewicz:2010:PSL:1807167.1807184}, where graph algorithms must be
executed as a sequence of iterations of computation and message passing.
Both those systems provide excellent performance, however they are not
declarative and have a steep learning curve.

@@

Linda~\cite{linda} is arguably the most famous coordination model. Linda
implements a data-driven coordination model and features a \emph{tuple space}
that can be manipulated using the following coordination directives:
\texttt{out(t)} to write a tuple \texttt{t} into the tuple space; \texttt{in(t)}
to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy
of the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a
process \texttt{p} in the tuple space and execute it in parallel.  Linda is be
implemented on top of many popular languages by simply creating a communication
and storage mechanism for the tuple space and then adding the directives as a
language library.

@@

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. But even more importantly, CLM differs from those
languages because coordination acts on data instead of processing units.
The abstraction is then raised by considering data and algorithmic aspects of
the program instead of focusing how processing units are used.
Furthermore, the CLM language is both a coordination language and a computation
language and there is no distinction between the two components.

\paragraph{Forward-Chaining Logic Programming}

Forward-chaining logic programming came into its own with the Datalog
language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
databases, it is now being increasingly used in different fields such as
distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a
Datalog-like language for programming distributed ensembles of modular robots.
Meld introduced the idea of sensing and action facts in order to sense and act
on the outside world, respectively. LCM uses sensing facts to sense the
underlying system that the program is running on and action facts act on that
system.

>>>>

\paragraph{Declarative Languages}
Programming paradigms such as \emph{logic} and \emph{functional}
programming have been extensively exploited for their implicit parallelism.
In logic languages such as Prolog, researchers took advantage of the
non-determinism of proof-search to evaluate subgoals in parallel with models
such as \emph{or-parallelism}~\cite{ali-86} and
\emph{and-parallelism}~\cite{Shen-92}.  In functional languages, the stateless
nature of computation allows multiple expressions to evaluate safely in
parallel. This has been explored in several languages such as
NESL~\cite{Blelloch:1996:PPA:227234.227246}, Id~\cite{Nikhil93anoverview}, and
more recently Data Parallel Haskell~\cite{nepal}.
NESL often got good performance, but was limited to a particular
application space. We also limit our application space, but also give the
programmer a declarative way to coordinate computation to improve the
performance of the program.

@@

Another interesting system is GraphLab~\cite{GraphLab2010}, a C++ framework for developing graph-based parallel machine
learning algorithms. GraphLab allows nodes to have read/write access to
different scopes through different concurrent access models in order to balance
performance and data consistency.  GraphLab also provides different schedulers
that dictate the order in which node's are computed, allowing the programmer to
optimize the program. Later in this paper, we will show how one GraphLab
scheduler can be implemented in CLM through the use of coordination facts.
Another, more restrictive, graph-based system is
Pregel~\cite{Malewicz:2010:PSL:1807167.1807184}, where graph algorithms must be
executed as a sequence of iterations of computation and message passing.
Both those systems provide excellent performance, however they are not
declarative and have a steep learning curve.

@@

Linda~\cite{linda} is arguably the most famous coordination model. Linda
implements a data-driven coordination model and features a \emph{tuple space}
that can be manipulated using the following coordination directives:
\texttt{out(t)} to write a tuple \texttt{t} into the tuple space; \texttt{in(t)}
to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy
of the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a
process \texttt{p} in the tuple space and execute it in parallel.  Linda was
implemented on top of many popular languages by simply creating a communication
and storage mechanism for the tuple space and then adding the directives as a
language library.

@@

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. CLM differs from those
languages because coordination acts on data instead of processing units.
The abstraction is then raised by considering data and algorithmic aspects of
the program instead of focusing on how processing units are used.
Furthermore, the CLM language is both a coordination language and a computation
language and there is no distinction between the two components.

\paragraph{Forward-Chaining Logic Programming}
This paradigm came into its own with the Datalog
language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
databases, it is now being increasingly used in different fields such as
distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a
Datalog-like language for programming distributed ensembles of modular robots.
Meld also introduced the idea of sensing and action facts in order to sense and act
on the outside world, respectively. LCM uses sensing facts to sense the
underlying system and action facts to act on that system.
